Evaluating an attempt to restore summer fire in the Northern Great Plains

### 2025-04-21 SpringSummerSeverity/paper/SummerFireRestorationEM1.Rmd
Spring/summer seasonality R1 update

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
author:
- Devan Allen McGranahan* and Jay P. Angerer
- Livestock and Range Research Laboratory, USDA Agricultural Research Service, Miles City, Montana, USA 
- "*Correspondence: Devan.McGranahan@usda.gov"
output: # html_document
  word_document:
    reference_docx: re_template.docx
    keep_md: yes
csl: restoration-ecology.csl
bibliography: 
  - SpringSummerFire.bib 
always_allow_html: true
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE)
pacman::p_load(tidyverse, knitr, kableExtra)
# Some caption tools 
  t_nums = captioner::captioner(prefix = "Table")
  f_nums = captioner::captioner(prefix = "Fig.")
  t_ref = function(x) { str_extract(t_nums(x), "[^:]*")   }
  f_ref = function(x) { str_extract(f_nums(x), "[^:]*")   }
```

@@

There is growing interest in diversifying human-managed fire regimes. 
In many North American grasslands, late growing season burns re-introduce fire to periods most prone to lightning-driven fire prior to wildfire suppression policies. 
We report here on restoring summer fire in central North Dakota, USA, from a research project in which summer fire was only successfully achieved in two out of four years. 
We use remotely-sensed imagery and local weather data to assess whether fuel or weather conditions limited burning in the summer, and to compare fire environmental conditions and subsequent burn severity across prescribed burns conducted in the spring and summer. 
Finally, we review historical data to determine if conditions have changed in either the spring or summer burn seasons over 42 years. 
We found that burn severity declined with fuelbed greenness but was independent of burn season---summer burns could effect as high of severity as spring burns despite having greener fuelbeds.
What little phenological change seems to have occurred at the study location---slightly greener fuelbeds and slightly lower relative humidity in spring---likely offset each other to some degree. 
Overall, we found little evidence that successful summer burns were anomalies, and conclude that it is reasonable for managers to incorporate late growing season fire into prescribed fire programs with the caveat that some summers will simply be too wet and/or too green to burn.

@@

@@

In this paper we report on mixed success in conducting summer prescribed fire in the Northern Great Plains. 
As researchers attempted to restore spatially-patchy prescribed fire to grazed mixed-grass prairie, two patterns of fire were introduced: one in which an entire one-quarter patch was burned each spring, and another in which half of the patch---one-eighth of the pasture---was burned in the spring, followed by the second half-patch scheduled for a late summer burn, in an attempt to provide a late-season flush of high-quality forage [e.g., @allred2011; @nichols2021] and determine whether a greater number of patches would increase spatial heterogeneity  [e.g., @mcgranahan2018b].
While all spring burns were completed as planned, summer fires were only successful in two out of four years. 
Here we present seasonal and long-term remotely-sensed burn severity and fuel data alongside weather information in considering whether successful summer burns are likely to be the exception or the norm based on historical patterns.

@@

A total of 40 prescribed fires were conducted over four years, 2017--2020. 
All 32 planned spring burns were completed, between 25 April and 25 May. 
Only eight of 16 planned summer burns were completed: 4 in 2017, and 4 in 2020, between 13-25 August. 
Burn units were surrounded with plowed mineral firebreaks and ignitions varied between backing and head fires, and interior strip ignitions, with handheld drip torches.

@@

Two types of data are presented here: remotely-sensed imagery from earth observation satellites, and weather data logged by the Center's mesonet weather station. 
We used remotely-sensed data from two separate sources. 
For the study period alone, 2017--2020, we used the European Space Agency's Sentinel-2 mission, which provides hyperspectral imagery at 10 $\times$ 10 m resolution at 5-10 day intervals. 
For historical context, 1990--2022, we used Landsat resources from the US Geological Survey, which provides hyperspectral imagery at 30 $\times$ 30 m resolution at 7-14 day intervals; Sentinel-2 only began in 2016, while the first Landsat mission launched in the mid 1980s.

Remotely-sensed data consisted of two indices: the Normalized Burn Ratio (NBR), which is used to estimate burn severity, and Normalized Difference Vegetation Index (NDVI), used to estimate greenness of vegetation. 
Sentinel-2 imagery was retrieved from the online Copernicus Browser using a custom EvalScript to return values for each index as a separate band in a geo-referenced raster file; each scene was inspected to ensure cloud-free imagery over pasture locations beginning with the first snow-free scene through to the final snow-free scene each year. 
To calculate burn severity, we subtracted the first post-fire image from the most-recent pre-fire image for each burn unit, generating a burn severity index known as differenced Normalized Burn Ratio (dNBR). 
Estimates of fuel greenness were made with the NDVI from the same most-recent pre-burn image.

Landsat imagery was processed with Google Earth Engine using a custom script to merge missions over time, mask clouds, and calculate NDVI for the burn seasons defined above, 1990--2022. 
See Supplemental Information for scripts used for both Copernicus and Google Earth Engine.

@@

Weather data---which included air temperature, relative humidity, dew point, windspeed, and vapor pressure deficit---were retrieved from the North Dakota Agricultural Weather Network's Streeter station [@ndawn2023]. 
To characterize day-of fire weather for successful burn days, we calculated mean values for operational burn periods (1000--1700) from hourly data for days recorded as having at least one completed fire. 
To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season. 
Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April- 31 October, 1990--2023).

@@

@@

Grassland fuelbeds were less green during each of the spring fire seasons (Sentinel-2 NDVI 0.3-0.4) than in the summer (NDVI > 0.5), with the two seasons in which summer burns could not be completed being the most green (NDVI > 0.6; `r f_ref("sentinel")`A). 
Burn severity ranged widely within and across seasons and years (`r f_ref("sentinel")`B). 
Among the 32 spring burns, there was a general decline in burn severity with increasing fuelbed greenness. 
The low number of successful summer burns precluded a linear assessment, but despite having categorically higher greenness, summer burning effected the same range of variability in burn severity as in the spring (`r f_ref("sentinel")`B).

Fuelbed greenness during successful burn seasons was close to the typical range over the 42-year historical period (`r f_ref("GreenRain")`A), while cumulative rainfall was at or below the typical historical range (`r f_ref("GreenRain")`B). 
Conversely, for years in which summer burns were not completed, both summer fuelbed greenness and accumulated rainfall by 1 August were atypically high relative to their historical ranges (`r f_ref("GreenRain")`).
<!-- Note that spring NDVI derived from Landsat (`r f_ref("GreenRain")`A) was slightly higher than the same data from Sentinel-2 (`r f_ref("sentinel")`A); the reason for the difference is unclear but might relate to the lower temporal resolution of Landsat imagery skewing the spring average with later, greener values and fewer early, snow-free images. -->
<!-- Notably, summer values appear quite consistent between Landsat and Sentinel-2.  -->

Observed fire weather varied widely within and across the four study years, and generally spanned the entire range of historical values, although wind speeds on the day of prescribed burns never exceeded the historical median in either season, and only one burn day exceeded the historical median relative humidity (`r f_ref("weather")`). 
Throughout the 42-year dataset, summer burn seasons generally had higher moisture, greater vapor pressure deficient, and lower wind speed than spring burn seasons.

Few substantial changes in the fire environment appear to have occurred since the 1990s in either spring or summer burn seasons (`r f_ref("historical")`). 
The two most notable changes---increased spring fuelbed greenness and lower spring dewpoint and relative humidity---might cancel each other out, with drier air potentially mitigating the impact of higher-moisture fine fuels. 
In two instances---spring windspeed and summer rainfall---the more apparent temporal trend was for greater variability around the typical range, rather than substantial shifts in typical values.

@@

We used remotely-sensed imagery and weather data to inform whether, in the four years summer burns were attempted in a research project in the Northern Great Plains, the two successful years, or the two unsuccessful years, were more likely outcomes based on typical fuel and weather conditions. 
Comparing seasonal values to historical ranges suggests that unsuccessful burn seasons were the anomalies. 
As such, successfully completing prescribed fire in the late growing season in the Northern Great Plains is a reasonable management objective. 
However, prairie fuelbeds might simply be too green to carry fire in some years, likely those with above-normal rainfall over the course of the growing season.

The data provide little indication that fire weather posed a barrier to successful summer burning. 
Overall, burn season fire weather was highly variable across study years, and burn day weather spanned the gamut (the exception was wind speed, which was below average in all study years). 
For example, for the four years summer burns were planned, one successful year and one unsuccessful year were at the high end of typical relative humidity, and the other successful and unsuccessful years were at the low end of typical relative humidity. 
In a study of these and similar grasslands that included a subset of the spring fires reported here, higher wind speed and greater vapor pressure deficit contributed to higher rates of spread, a measure of fire intensity [@mcgranahan2023a]. 
But overall, the statistical analysis provided little explanation for the substantial variability in fire behavior.

<!-- Experience does suggest fire weather conditions can limit specific prescribed fire operations.  -->
<!-- For example, a burn was planned around a favorable weather forecast on August 13, 2018, and a full crew was on-site that morning.  -->
<!-- But relative humidity did not fall as predicted in the spot weather forecast, remaining above 50\% until 1500 (the burn was called off by 1400 after repeated test fires failed to spread).  -->
<!-- Certainly, actual weather defies forecasts in all seasons.  -->
<!-- But in the context of summer prescribed fire, when fuels are more green and windspeed generally lower (`r f_ref("historical")`), stubborn relative humidity is more of a barrier in the absence of mitigating factors.  -->

Although trends towards lower relative humidity might mitigate trends towards greener fuelbeds in the spring, cumulative seasonal rainfall, fuelbed greenness, and relative humidity likely all increase together in the summer, when plants are photosynthetically active. 
Conversely, our historical comparisons suggest the spring burn season is perhaps most susceptible to global changes, although complex interactions between drivers of global change introduce substantial uncertainty. 
In addition to continental-scale increases in air temperature and variability in precipitation [@kukal2016; @pendergrass2017], fire managers in the Northern Great Plains must also deal with local changes. 
For example, native grasslands of the Northern Plains---including the pastures considered here---have been invaded with exotic, cool-season grasses that green up early, change fuel structure, and reduce rate of spread [@gasch2020; @yurkonis2019].

@@

Research into opportunities to conduct prescribed fire might consider phenological changes in the fire environment, rather than conventional attempts to assess changes in the availability of "weather windows"---the set of conditions under which prescribed fire might possibly be conducted [e.g., @weir2011]. 
While some research has used the concept of the weather window to identify seasonal shifts in  potentially favorable burn conditions---e.g., @yurkonis2019 showed a decrease in early spring burn days and an increase in late summer burn days for central North Dakota---the focus on prescriptions generally masks interactions between mitigating and exacerbating factors, fails to account for fuelbed dynamics, and is agnostic to antecedent weather and fuel conditioning. 
We suggest instead an approach that uses the type of data presented here to classify the seasonality of fuels and fire weather to investigate temporal changes in the phenology of the fire environment as a whole.

@@

```{r sentinel, fig.cap = sent_cap}
sent_cap <- f_nums(name = "sentinel", 
                   caption = "\\label{sentinel}Sentinel imagery summarized for the study years. A: Normalized Difference Vegetation Index (NDVI) from four adjacent unburned pastures avoids any confounding effects of prescribed burns. X denotes seasons in which burns were not completed. B: Burn severity (differenced Normalized Burn Ratio, dNBR) plotted against fuelbed greenness (NDVI) retrieved from the pre-burn Sentinel-2 image used for pre-fire NBR. Points represent mean for individual prescribed fires with associated standard error; circles and solid trendline indicate spring burns while broken trendline and triangles denote summer burns.")
include_graphics('ndvi_gg-1.png')
```

@@

```{r historical, fig.cap = hist_cap}
hist_cap <- f_nums(name = "historical", 
                   caption = "\\label{historical}Temporal trends in the phenology of the fire environment. Boxplots represent data from the first decade of the dataset used here (1990-1999) to the final decade (2012-2022), by burn season. Horizontal lines represent average conditions for the four study years (2017-2020).")
include_graphics('historical_phenology-1.png')
```

>>>>

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
#subtitle: 'Running head: Summer burning in the Northern Plains'
author:
- Devan Allen McGranahan* and Jay P. Angerer
- Livestock and Range Research Laboratory, USDA Agricultural Research Service, Miles City, Montana, USA 
- "*Correspondence: Devan.McGranahan@usda.gov"
output: # html_document
  word_document:
    reference_docx: re_template.docx
    keep_md: yes
csl: environmental-management.csl
bibliography: 
  - SpringSummerFire.bib 
always_allow_html: true
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE)
pacman::p_load(tidyverse, knitr, kableExtra, pander)
# Some caption tools 
  t_nums = captioner::captioner(prefix = "Table")
  f_nums = captioner::captioner(prefix = "Fig.")
  t_ref = function(x) { str_extract(t_nums(x), "[^:]*")   }
  f_ref = function(x) { str_extract(f_nums(x), "[^:]*")   }

load('../data/z_tab.Rdata')
load('../data/comp_tab.Rdata')
```

**Author contributions:** Conception, analysis, writing: DAM; Programming and editing: JPA

@@

There is growing interest in diversifying human-managed fire regimes. 
In many North American grasslands, late growing season burns re-introduce fire to periods most prone to lightning-driven fire prior to wildfire suppression policies. 
We report here on restoring summer fire in central North Dakota, USA, from a research project in which summer burns were only completed in two out of four years for which summer burns were planned. 
We use remotely-sensed imagery and local weather data to assess whether fuel or weather conditions limited burning in the summer, and to compare fire environmental conditions and subsequent burn severity across prescribed burns conducted in the spring and summer. 
Finally, we review historical data to determine if conditions have changed in either the spring or summer burn seasons over 42 years. 
Although burn severity generally declined with fuelbed greenness in the spring, summer burns could effect as high of severity as spring burns despite having greener fuelbeds.
What little phenological change seems to have occurred at the study location over 42 years---slightly greener fuelbeds and slightly lower relative humidity in spring---likely offset each other to some degree. 
Overall, we found little evidence that being able to complete summer burns was anomalous, and conclude that it is reasonable for managers to incorporate late growing season fire into prescribed fire programs with the caveat that some summers will simply be too wet and/or too green to burn.

@@

**Statements and Declarations** The authors declare no competing interests.

@@

In this paper we report on mixed results in conducting summer prescribed fire in the Northern Great Plains. 
As researchers attempted to restore spatially-patchy prescribed fire to grazed mixed-grass prairie, two patterns of fire were introduced: one in which an entire one-quarter patch was burned each spring, and another in which half of the patch---one-eighth of the pasture---was burned in the spring, followed by the second half-patch scheduled for a late summer burn, in an attempt to provide a late-season flush of high-quality forage [e.g., @allred2011; @nichols2021] and determine whether a greater number of patches would increase spatial heterogeneity  [e.g., @mcgranahan2018b].
While all spring burns were completed as planned, summer fires were only completed in two out of four years. 
Here we present seasonal and long-term remotely-sensed burn severity and fuel data alongside weather information in considering whether being able to conduct summer burns was exceptional or more likely to be the norm based on historical patterns.

@@

Forty prescribed fires were conducted over four years, 2017--2020. 
All 32 planned spring burns were completed, between 25 April and 25 May. 
Only eight of 16 planned summer burns were completed: 4 in 2017, and 4 in 2020, between 13-25 August. 
Burn units were surrounded with plowed mineral firebreaks and ignitions varied between backing and head fires, and interior strip ignitions, with handheld drip torches.

@@

Two types of data are presented here: remotely-sensed imagery from earth observation satellites, and weather data logged by the Center's mesonet weather station. 
We used remotely-sensed data from two separate sources. 
For the study period alone, 2017--2020, we used the European Space Agency's Sentinel-2 mission, multispectral imagery at 10 $\times$ 10 m resolution in the visible and near-infrared region and 20 $\times$ 20 m in the red-edge and shortwave infrared regions.
Data are collected and produced at 5-10 day intervals. 
For historical context, 1990--2022, we used Landsat resources from the US Geological Survey (Landsat 5, 7, and 8 Level 2, Collection 2, Tier 1 Surface Reflectance), which provide multispectral imagery at 30 $\times$ 30 m resolution at 7-14 day intervals; Sentinel-2 only began in 2016, while the first Landsat 5 mission launched in the mid 1980s.

Remotely-sensed data consisted of two indices: the Normalized Burn Ratio (NBR; @key2006), which is used to estimate burn severity, and Normalized Difference Vegetation Index (NDVI; @rousejr1974), used to estimate greenness of vegetation. 
Sentinel-2 imagery was retrieved from the online Copernicus Browser using a custom EvalScript to return values for each index as a separate band in a geo-referenced raster file; each scene was inspected to ensure cloud-free imagery over pasture locations beginning with the first snow-free scene through to the final snow-free scene each year. 
To calculate burn severity, we subtracted the first post-fire image from the most-recent pre-fire image for each burn unit, generating a burn severity index known as differenced Normalized Burn Ratio (dNBR; @key2006). 
Estimates of fuel greenness were made with the NDVI from the same most-recent pre-burn image.

Landsat imagery was processed with Google Earth Engine [@gorelick2017] using a custom script to merge missions over time, mask clouds, and calculate NDVI for the burn seasons defined above, 1990--2022. 
See Supplemental Information for scripts used for both Copernicus and Google Earth Engine.

@@

Weather data---which included air temperature, relative humidity, dew point, windspeed, and vapor pressure deficit---were retrieved from the North Dakota Agricultural Weather Network's Streeter station [@ndawn2023]. 
To characterize day-of fire weather for burn days, we calculated mean values for operational burn periods (1000--1700 hrs) from hourly data for days recorded as having at least one completed fire. 
To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season. 
Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April--31 October, 1990--2023). 
Historical trends in both Landsat NDVI and fire weather are available in Supplemental Information.

@@

## Statistical analysis

Difference in NDVI between spring and summer for the four study years was tested via analysis of deviance ($\chi^2$ test), in which with a linear mixed-effect model was compared against an intercept-only, null model, each fit with the `lmer` function in the *lme4* package for **R** [@bates2015]. 
Linear relationships between dNBR and NDVI were first tested for a difference between spring and summer trends using analysis of deviance ($\chi^2$ test), then within each season ($t$ test), using generalized mixed-effect regression models fit with a Gamma distribution using `glmer` from *lme4*. 
Pasture (block) and year were fit as random effects.

To compare mean values for each year of the study period to historical ranges, we calculated $z$ scores as a measure of deviation from the mean [@schober2021] by subtracting the historical mean from the seasonal value, and dividing by the historical standard deviation. 
*Anomalies* were defined as values more than one standard deviation above or below the historical mean; *statistical significance* at $\alpha$ = 0.05 occurs when $z > |2|$.

Potential temporal changes in fuel greenness and fire weather variables from early in the historical dataset (1990--1999) and recent values (2012-2022) were tested as differences in means using ordinary least squares regression ($t$ test) at $\alpha$ = 0.05.

@@

Grassland fuelbeds were less green during each of the spring fire seasons (Sentinel-2 NDVI 0.3-0.4) than in the summer (NDVI > 0.5; $\chi^2$ = 134, $P <$ 0.001), with the two seasons in which summer burns could not be completed being the most green (NDVI > 0.6; `r f_ref("sentinel")`A). 
Burn severity ranged widely within and across seasons and years (`r f_ref("sentinel")`B); despite having categorically higher greenness (`r f_ref("sentinel")`A), summer burning effected the same range of variability in burn severity as in the spring.
The relationship between dNBR and NDVI varied between seasons ($\chi^2$ = 8.4, $P <$ 0.01). 
Among summer burns, there was no linear relationship between dNBR and NDVI ($t$ = -0.97, $P >$ 0.05).
Among spring burns, there was a general decline in burn severity with increasing fuelbed greenness ($t$ = -4.7, $P <$ 0.001).

Fuelbed greenness, as measured by Landsat NDVI, during the four spring burn seasons was within or slightly above typical values for the 42-year historical period (`r f_ref("GreenRain")`A; `r t_ref("GreenRain_z")`).
Summer seasons with completed burns were very near the historical mean NDVI (`r t_ref("GreenRain_z")`), while summer seasons without completed burns had the highest NDVI anomalies (2018, $z$ = 0.76; 2019, $z$ = 1.52). 
In terms of cumulative rainfall, all four spring seasons were at or below historical values (`r f_ref("GreenRain")`B; `r t_ref("GreenRain_z")`). 
Years in which summer burns were not completed had the highest anomalies in cumulative precipitation (2018, $z$ = 1.6; 2019, $z$ = 0.8). 
Conversely, cumulative precipitation in years in which summer burns were completed was at least 1 standard deviation below the historical mean (2017, $z$ = -1.04; 2020, $z$ = -1.3).
<!-- Note that spring NDVI derived from Landsat (`r f_ref("GreenRain")`A) was slightly higher than the same data from Sentinel-2 (`r f_ref("sentinel")`A); the reason for the difference is unclear but might relate to the lower temporal resolution of Landsat imagery skewing the spring average with later, greener values and fewer early, snow-free images. -->
<!-- Notably, summer values appear quite consistent between Landsat and Sentinel-2.  -->

Day-of-burn fire weather varied widely within and across the four study years, and generally spanned the entire range of historical values (`r f_ref("weather")`). 
However, wind speeds on the day of prescribed burns never exceeded the historical median in either season, and only one burn day exceeded the historical median relative humidity (`r f_ref("weather")`). 
Seasonal averages in fire weather for each of the four study years also ranged widely relative to historical means. 
In summer, especially, there were no patterns in the magnitude and signs of fire weather anomalies that distinguished years in which summer burns were completed, and those in which they were not (`r t_ref("weather_z")`).

Few substantial changes in the fire environment appear to have occurred since the 1990s in either spring or summer burn seasons (`r f_ref("historical")`, `r t_ref("comp_stats")`). 
The only statistically-significant change was an increase in spring fuelbed greenness ($t$ = 3.8, $P$ < 0.001).
In two instances---spring windspeed and summer rainfall---the more apparent temporal trend was for greater variability around the typical range, rather than substantial shifts in typical values.

@@

We used remotely-sensed imagery and weather data to assess whether, in the four years summer burns were planned in a research project in the Northern Great Plains, the fuel and weather conditions of the two years in which burns were completed, or conditions of the two years in which burns were not completed, were more likely to represent normal conditions based on historical patterns. 
Comparing seasonal values to historical ranges suggested that not being able to complete summer burns were the anomalies. 
As such, completing prescribed fire in the late growing season in the Northern Great Plains is a reasonable management objective. 
However, prairie fuelbeds might simply be too green to carry fire in some years, likely those with above-normal rainfall over the course of the growing season.

At a seasonal level, these data provide little indication that fire weather posed a barrier to conducting summer fire, and instead pointed to fuelbed greenness.
This pattern complements regional wildfire trends, which show an increase in wildfire frequency with lower summer precipitation and higher temperatures and drought indices [@clabo2018]. 
Overall, burn season fire weather in our data was highly variable across study years, and burn day weather spanned the gamut (the exception was wind speed, which was below average in all study years). 
Take, for example, relative humidity: for the four years summer burns were planned, two years were above average ($z$ scores 0.9 and 0.5), and two years were below average ($z$ scores -0.6 and -0.9), but each set included a year in which summer burns were completed and a year in which burns were not completed. 
In a study of these and similar grasslands that included a subset of the spring fires reported here, higher wind speed and greater vapor pressure deficit contributed to higher rates of spread, a measure of fire intensity [@mcgranahan2023a]. 
But overall, the statistical analysis provided little explanation for the substantial variability in fire behavior.

However, experience does suggest fire weather conditions can limit specific prescribed fire operations.
For example, a burn was planned around a favorable weather forecast on 13 August 2018, and a full crew was on-site that morning. 
But relative humidity did not fall as predicted in the spot weather forecast, remaining above 50\% until 1500 hrs (the burn was called off by 1400 hrs after repeated test fires failed to spread).
Certainly, actual weather defies forecasts in all seasons.
But in the context of summer prescribed fire, when fuels are more green and windspeed potentially lower (`r f_ref("historical")`), stubborn relative humidity is more of a barrier in the absence of mitigating factors.

Consistent with a wide body of research on land surface phenology, the single statistically-significant change in the fire environment over the 42 year period at our study location was increased spring fuelbed greenness (`r t_ref("comp_stats")`).
Earlier spring green-up over recent decades has been documented worldwide in northern temperate ecosystems generally [@xu2018; @ren2024], and in temperate US rangelands, specifically [@zimmer2022]. 
While earlier spring green-up in grasslands has been broadly attributed to increased spring precipitation and higher temperatures [@post2022], managers in the Northern Great Plains must also deal with non-native cool-season grasses that are particularly responsive to earlier onset of the growing season [@gasch2020]. 
Conversely, observed trends toward lower spring dewpoint and relative humidity (`r f_ref("historical")`), although not statistically-significantly different at this single weather station, suggests drier air could potentially mitigate spring fuelbed greenness.

While these historical comparisons suggest the spring burn season is perhaps most susceptible to global changes, complex interactions between drivers of global change introduce substantial uncertainty.
In addition to continental-scale increases in air temperature and variability in precipitation [@kukal2016; @pendergrass2017], fire managers in the Northern Great Plains must also deal with local changes. 
For example, native grasslands in the region---including the rangeland considered here---have been invaded with exotic, cool-season grasses that green up early, change fuel structure, and reduce rate of spread [@yurkonis2019].
Even if trends towards lower relative humidity might mitigate greener fuelbeds in the spring, cumulative seasonal rainfall, fuelbed greenness, and relative humidity likely all increase together in the summer, when plants are photosynthetically active, creating potential barriers to summer burns.
Conversely, the non-native cool-season grasses of the Northern Great Plains, especially *Poa pratensis*, are widely known to be susceptible to summer drought [@etter1951; @meyer1989; @liu2008];  dormancy due to water stress reduces fine fuel moisture and potentially facilitates fire spread. 
Given these bimodalities, we suggest that the greater the presence of non-native cool-season grasses in fuelbeds of the Northern Great Plains, the more extreme managers might expect the potential magnitude of fuel and fire responses to year-to-year variability in precipitation.

@@

Research into opportunities to conduct prescribed fire might apply the "land surface phenology" approach to the fire environment [@hanes2014; @zimmer2022], rather than conventional attempts to assess changes in the availability of "weather windows"---the set of conditions under which prescribed fire might possibly be conducted [e.g., @weir2011]. 
While some research has used the concept of the weather window to identify seasonal shifts in  potentially favorable burn conditions---e.g., @yurkonis2019 showed a decrease in early spring burn days and an increase in late summer burn days for central North Dakota---the focus on prescriptions generally masks interactions between mitigating and exacerbating factors, fails to account for fuelbed dynamics, and is agnostic to antecedent weather and fuel conditioning. 
We suggest instead an approach that uses the type of data presented here to classify the seasonality of fuels and fire weather to investigate temporal changes in the phenology of the fire environment as a whole.

@@

```{r tab_caps}
gr_tab_cap <- t_nums(name = 'GreenRain_z', 
                       caption = '\\label{GreenRain_z}Seasonal deviations from historical averages in Normalized Differenced Vegetation Index (NDVI) and year-to-date precipitation for spring and summer burn seasons, as presented in Fig. 2. Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the historical mean. 2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years. ')
wx_tab_cap <- t_nums(name = 'weather_z', 
                       caption = '\\label{weather_z}Seasonal deviations from historical averages in four fire weather variables for spring and summer burn seasons, as presented in Fig. 3. Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the historical mean. 2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years ')
hist_tab_cap <- t_nums(name = 'historical_z', 
                       caption = '\\label{historical_z}Seasonal deviations in fuel conditions and fire weather for spring and summer burn seasons, relative to two decades (1990-1999, 2012-2022) as presented in Fig. 4. Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the mean for the given historical range. 2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years')
comp_tab_cap <- t_nums(name = 'comp_stats', 
                       caption = '\\label{comp_stats}Results of statistical models comparing, within each season, early period (1990-1999) to late period (2012-2022) values for fuelbed greenness and five fire weather variables in the historical dataset (Fig. 4).')
```

```{r GreenRain_z, eval = TRUE, tab.cap=gr_tab_cap}
 z_tab$ndvi_prcp %>%
  kable() 
```

```{r weather_z, eval = TRUE, tab.cap=wx_tab_cap}
 z_tab$wxd %>%
  kable() 
```

```{r comp_stats, eval = TRUE, tab.cap=comp_tab_cap}
 comp_tab %>%
  kable() 
```

```{r historical_z, eval = FALSE, tab.cap=hist_tab_cap}
 z_tab$hist %>%
  kable() 
```

```{r sentinel, fig.cap = sent_cap}
sent_cap <- f_nums(name = "sentinel", 
                   caption = "\\label{sentinel}Sentinel imagery summarized for the study years. A: Normalized Difference Vegetation Index (NDVI), gathered from four adjacent unburned pastures to avoid confounding effects of prescribed burns, was greater in summer burn seasons than spring burn seasons ($\\chi^2$ = 134, $P <$ 0.001). X denotes seasons in which burns were not completed. B: Burn severity (differenced Normalized Burn Ratio, dNBR) plotted against fuelbed greenness (NDVI) retrieved from the pre-burn Sentinel-2 image used for pre-fire NBR. Points represent mean for individual prescribed fires with associated standard error: circles indicate spring burns while triangles denote summer burns. The statistical relationship between dNBR and NDVI differs between seasons in generalized mixed-effect models ($\\chi^2$ = 8.4, $P$ < 0.01). dNBR declined as NDVI increased in spring burns ($t$ = -4.7, $P$ < 0.001; solid trendline, plotted as ordinary least-squares regression with 95% confidence interval, $R^2$ = 0.22); summer burns have no significant relationship (broken trendline). ")
include_graphics('ndvi_gg-1.png')
```

@@

```{r historical, fig.cap = hist_cap}
hist_cap <- f_nums(name = "historical", 
                   caption = "\\label{historical}Temporal comparison of phenological changes in the fire environment over 42 years. Boxplots represent data from the first decade of the dataset used here (1990-1999) to the final decade (2012-2022), by burn season. Horizontal lines represent average conditions for the four study years (2017-2020). Table 3 reports statistical tests of early vs. late periods for each season by each variable; Supplemental Information Table 1 reports z scores for each study year against historical values.")
include_graphics('historical_phenology-1.png')
```

### 2025-04-21 SpringSummerSeverity/paper/SummerFireRestorationRE0.Rmd
Spring/summer seasonality R1 update

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
author:
- Devan Allen McGranahan* and Jay P. Angerer
- Livestock and Range Research Laboratory, USDA Agricultural Research Service, Miles City, Montana, USA 
- "*Correspondence: Devan.McGranahan@usda.gov"
output: # html_document
  word_document:
    reference_docx: re_template.docx
    keep_md: yes
csl: restoration-ecology.csl
bibliography: 
  - SpringSummerFire.bib 
always_allow_html: true
---

@@

# Abstract

@@

@@

Weather data---which included air temperature, relative humidity, dew point, windspeed, and vapor pressure deficit---were retrieved from the North Dakota Agricultural Weather Network's Streeter station [@ndawn2023]. 
To characterize day-of fire weather for successful burn days, we calculated mean values for operational burn periods (1000--1700) from hourly data for days recorded as having at least one completed fire. 
To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season. 
Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April- 31 October, 1990--2023).

@@

>>>>

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
subtitle: 'Running head: Summer burning in the Northern Plains'
author:
- Devan Allen McGranahan* and Jay P. Angerer
- Livestock and Range Research Laboratory, USDA Agricultural Research Service, Miles City, Montana, USA 
- "*Correspondence: Devan.McGranahan@usda.gov"
output: # html_document
  word_document:
    reference_docx: re_template.docx
    keep_md: yes
csl: restoration-ecology.csl
bibliography: 
  - SpringSummerFire.bib 
always_allow_html: true
---

@@

**Author contributions:** Conception, analysis, writing: DAM; Programming and editing: JPA

\newpage 
# Abstract

@@

# Implications for practice

* Successful Rx fire in the summer is likely the norm, not the exception. 
* At least at this specific research station, there is little evidence that fuel and weather conditions have changed over 42 years except for slightly greener spring fuels and lower relative humidity, which might cancel each other out.

@@

Weather data---which included air temperature, relative humidity, dew point, windspeed, and vapor pressure deficit---were retrieved from the North Dakota Agricultural Weather Network's Streeter station [@ndawn2023]. 
To characterize day-of fire weather for successful burn days, we calculated mean values for operational burn periods (1000--1700) from hourly data for days recorded as having at least one completed fire. 
To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season. 
Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April--31 October, 1990--2023).

@@

The authors declare no conflict of interest.
